Non-word repetition in Yélî Dnye

### 2020-12-21 manuscript.tex
add indiv diffs disc subsection draft

Previous work typically steers clear of articulartorily and/or
acoustically challenging sounds, but we included some in our experiment
to more adequately represent Yélî Dnye's phonology. We ultimately used a
relatively large number of items that would enable us to explore both
variation in structural complexity and in more vs.~less challenging
sounds. However, aware that this large item inventory might render the
task longer and more tiresome, we split items across children (see
below). Naturally, designing the task in this way may make the study of
individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences in NWR are relatively small, and would not be
detectable with the sample size that we could collect during our short
visit. That said, we contribute to the literature by also reporting
descriptive analyses of individual variation that could potentially be
integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age (Farmani
et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
Stackhouse, \& Wells, 2005), and differences relating to bilingual
language exposure (Brandeker \& Thordardottir, 2015; Meir \&
Armon-Lotem, 2017; Meir et al., 2016). Previous work typically finds no
significant differences as a function of maternal education (e.g.,
Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or
child gender (Chiat \& Roy, 2007). Although past research has not often
investigated potential effects of birth order on NWR, there is a sizable
literature on these effects in other language tasks (Havron et al.,
2019), and therefore we report on these too.

@@

\begin{itemize}
\tightlist
\item
  Does the cross-linguistic frequency of sounds in the stimuli predict
  NWR scores? Are rarer sounds more often substituted by commoner
  sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation in NWR scores attributable to child age, sex,
  birth order, monolingual status, and/or parental education?
\end{itemize}

@@

We investigated the effect of item complexity on NWR scores by varying
both the number of syllables in the item and its average segmental
frequency. Based on previous work, we had predicted that children would
have higher NWR scores for shorter items. That said, previous work has
shown both very small (Piazzalunga et al., 2019) and very large (Cristia
et al., 2020; Jaber-Awida, 2018) effects of stimulus length and,
further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic
words are nearly equally frequent in the current language, with
trisyllabic and longer words making up a non-trivial 10\% of the
remaining words. Compare this to, for example, English, which is
substantially more skewed toward monosyllabic word forms **Alex I'm
going off your note here (``Prediction for Yélî made before seeing the
data: The length distribution in Yélî words is more balanced than that
in English, and thus the score decline for poly- versus mono-syllables
may be less pronounced than that for English.''``). I don't have a
reference for this, can you please finish the thought or nix this
bit?**. Setting aside our monosyllabic stimuli, which all contained
typologically infrequent segments, we can examine effects of item length
among the remaining stimuli, which range between 2 and 4 syllables long.
While indeed NWR scores were overall lower for longer items (e.g., see
Figure \ref{fig:prevlit-fig}), the effect of item length was not
significant in a statistical model that additionally accounted for age
and random effects of item and participant. In light of mixed prior
results of item length, we propose two possible (and non-mutually
exclusive) explanations for this minimal impact of item length. First,
further extensions of this type of analysis in more populations may
reveal that, in general (and cross-linguistically), item length effects
are variable between languages, potentially reflecting the distribution
of word lengths in the ambient language and other (morpho-)phonological
tendencies in the lexicon. Second, above and beyond these
language-specific effects, the general impact of item length on NWR
score may be relatively small, as shown in Piazzalunga et al.'s (2019)
study on Italian and as borne out in the current dataset once
controlling for other factors.

@@

\begin{itemize}
\tightlist
\item
  Is individual variation explainable by child age, sex, birth order,
  monolingual status, and/or parental education?
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\item
  Children's scores increases with child age.
\item
  General prediction: previous NWR evidence on this is mixed, but
  general findings on language development suggest that children whose
  mothers are more educated score higher in standardized language tests
  (eg ppvt) than children whose mothers are less educated.
\item
  Prediction made before seeing data: In the Rossel community, formal
  education plays an extremely minor role in ensuring individual's
  success, is not a good index of relative socio-economic status, and
  furthermore there is only a narrow range of variation in maternal
  educational attainment. This may lead to no or only very small
  advantages for children whose mothers are more educated, provided that
  the causal chain between maternal education and child language is via
  SES more broadly. However, if education directly boosts maternal
  verbal skills and the incidence of verbal behavior (as suggested by
  CITE), then we should still see a difference along this factor.
\item
  We did not find this
\item
  General prediction: To our knowledge, there is no previous NWR work on
  this, but other research suggests that first-born children score
  higher in standardized language tests later-born children.
\item
  Prediction before seeing data: One main causal path between birth
  order and language development is via parental input (CITE). Given our
  arguments above for how mothers may not be as important among Rossel
  people than in other places, then the difference in scores between
  first borns and later borns may be small
\item
  we did not find a sig effect, but this is a small effect (eg d=.2 in
  Havron et al)
\end{enumerate}

>>>>

Previous work typically steers clear of articulatorily and/or
acoustically challenging sounds, but we included some in our experiment
to more adequately represent Yélî Dnye's phonology. We ultimately used a
relatively large number of items that would enable us to explore both
variation in structural complexity and in more vs.~less challenging
sounds. However, aware that this large item inventory might render the
task longer and more tiresome, we split items across children (see
below). Naturally, designing the task in this way may make the study of
individual variation within the population more difficult because
different children are exposed to different items. However, a review of
previous work on individual variation suggested to us that effects of
individual differences in NWR are relatively small, and would not be
detectable with the sample size that we could collect during our short
visit. That said, we contribute to the literature by also reporting
descriptive analyses of individual variation that could potentially be
integrated in meta-analytic efforts.

Based on previous work, we looked at potential changes with age (Farmani
et al., 2018; Kalnak, Peyrard-Janvid, Forssberg, \& Sahlén, 2014; Vance,
Stackhouse, \& Wells, 2005). Previous work typically finds no
significant differences as a function of maternal education (e.g.,
Farmani et al., 2018; Kalnak et al., 2014; Meir \& Armon-Lotem, 2017) or
child gender (Chiat \& Roy, 2007). Although past research has not often
investigated potential effects of birth order on NWR, there is a sizable
literature on these effects in other language tasks (Havron et al.,
2019), and therefore we report on these too.

@@

\begin{itemize}
\tightlist
\item
  Does the cross-linguistic frequency of sounds in the stimuli predict
  NWR scores? Are rarer sounds more often substituted by commoner
  sounds?
\item
  How does score change as a function of item length in number of
  syllables?
\item
  Is individual variation in NWR scores attributable to child age, sex,
  birth order, and/or maternal education?
\end{itemize}

@@

We investigated the effect of item complexity on NWR scores by varying
both the number of syllables in the item and its average segmental
frequency. Based on previous work, we had predicted that children would
have higher NWR scores for shorter items. That said, previous work has
shown both very small (Piazzalunga et al., 2019) and very large (Cristia
et al., 2020; Jaber-Awida, 2018) effects of stimulus length and,
further, the Yélî Dnye dictionary suggests that mono- and bi-syllabic
words are nearly equally frequent in the current language, with
trisyllabic and longer words making up a non-trivial 10\% of the
remaining words. Compare this to, for example, English, which is
substantially more skewed toward monosyllabic word forms **M2A: Alex I'm
going off your note here (``Prediction for Yélî made before seeing the
data: The length distribution in Yélî words is more balanced than that
in English, and thus the score decline for poly- versus mono-syllables
may be less pronounced than that for English.''``). I don't have a
reference for this, can you please finish the thought or nix this
bit?**. Setting aside our monosyllabic stimuli, which all contained
typologically infrequent segments, we can examine effects of item length
among the remaining stimuli, which range between 2 and 4 syllables long.
While indeed NWR scores were overall lower for longer items (e.g., see
Figure \ref{fig:prevlit-fig}), the effect of item length was not
significant in a statistical model that additionally accounted for age
and random effects of item and participant. In light of mixed prior
results of item length, we propose two possible (and non-mutually
exclusive) explanations for this minimal impact of item length. First,
further extensions of this type of analysis in more populations may
reveal that, in general (and cross-linguistically), item length effects
are variable between languages, potentially reflecting the distribution
of word lengths in the ambient language and other (morpho-)phonological
tendencies in the lexicon. Second, above and beyond these
language-specific effects, the general impact of item length on NWR
score may be relatively small, as shown in Piazzalunga et al.'s (2019)
study on Italian and as borne out in the current dataset once
controlling for other factors.

@@

A review of previous work (see Introduction) suggested that our
anticipated sample size would not be sufficient to detect most
individual differences using NWR. We give a brief overview of individual
difference patterns of four types in the present data---age, sex, birth
order, and maternal education---hoping that these findings can
contribute to future meta-analytic efforts aggregating over smaller
studies such as ours.

Following prior work, we expected that NWR scores would increase with
participant age (Farmani et al., 2018; Kalnak et al., 2014; Vance et
al., 2005). Indeed, age was significantly correlated with NWR score and
also showed up as a significant predictor of NWR score when included as
a control factor in the analyses of both item length and average
segmental frequency. In brief, our results underscore the idea that
phoonlogical development continues well past the first few years of
life, extending into middle childhood and perhaps later (Hazan \&
Barrett, 2000).

In contrast, previous work shows little evidence for effects of maternal
education (e.g., Farmani et al., 2018; Kalnak et al., 2014; Meir \&
Armon-Lotem, 2017) or participant gender (Chiat \& Roy, 2007) on NWR
scores. In addition to this prior work, education on Rossel Island,
while generally highly valued, is not at all essential to ensuring one's
success in society and may not be a reliable index of local
socioeconomic variation. There is also limited variation in maternal
education across the families in the region of the island where we
sampled. We therefore expected little evidence for impact of either
participant gender or maternal education in the present study. On the
other hand, these predictors have established effects on other language
development measures (\textbf{REFS: M2A: Alex go ahead and pick your
faves here}). So to the extent that NWR scores share causal links to
gender-based differences ein development and maternal linguistic input
with these other language ouctome measures, we might then expect these
factors to appear in NWR data. In fact, neither participant gender nor
maternal education were correlated with NWR score in the current data.

Last but not least, we investigated whether birth order might affect NWR
scores, as it does other language tasks, resulting in first-born
children showing higher scores on standardized language tests than
later-born children (Havron et al., 2019), presumably because later-born
children receive a smaller share of maternal input than their older
siblings. Given shared caregiving practices and the hamlet organization
typical of Rossel communities, children have many sources of adult and
older child input that they encounter on a daily basis and first born
children quickly integrate with a much larger pool of both older and
younger children with whom they partly share caregivers. Therefore we
expected that any effects of birth order on NWR would be attenuated in
this context. In line with this prediction, our descriptive analysis
showed no correlation between birth order and NWR score.
